---
title: "PS3"
output: pdf_document
date: "2025-02-25"
---

```{r}
library(dplyr)
library(ggplot2)
library(fixest)
library(tidyr)
```


Q1
```{r}
###################################### Data Generation
set.seed(1234)

n_units <- 1000

#50 states
states <- 1:50
unit_states <- sample(states, n_units, replace = TRUE) ## assign 1-1000 to different states

#Policy years
g_list <- c(1996, 2002, 2008, 2014)
state_groups <- sample(g_list, 50, replace = TRUE)  # every state is assigned to a policy year
unit_groups <- state_groups[unit_states]

#years
years <- 1990:2020
n_years <- length(years)

#construct the panel data
data <- expand.grid(unit = 1:n_units, year = years)
data$state <- rep(unit_states, each = n_years)  
data$G <- rep(unit_groups, each = n_years)  # rep: replicate the element n times

#unit fixed effect α_i
alpha <- rnorm(n_units, mean = data$state / 1, sd = 1)
data$alpha <- rep(alpha, each = n_years)

#time fixed effect λ_t
data$lambda <- NA 
time_FE_errors <- rnorm(1, mean = 0, sd = 1)

for (g in g_list) {
  indices <- which(data$G == g)  ## the number of rows that G_i=g
  data$lambda[indices] <- 0.1 * (data$year[indices] - g) + time_FE_errors
}


#idiosyncratic error ε_it
data$epsilon <- rnorm(n_units * n_years, mean = 0, sd = 0.5)

#treatment effect τ_it
data$treatment_effect <- (data$year - data$G + 1) * (data$year >= data$G)

# Y_it
data$Y <- (2020 - data$G) + data$alpha + data$lambda + data$treatment_effect + data$epsilon

head(data)
```


```{r}
## Plot ATT
## ATT
ATT_data <- data %>%
  group_by(year, G) %>%
  summarise(ATT = mean(treatment_effect), .groups = 'drop')


## plot ATT & year t
ggplot(ATT_data, aes(x = year, y = ATT, color = factor(G))) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "ATT over t",
       x = "t",
       y = "ATT",
       color = "Policy Enactment Year")


## plot ATT & t-g
ggplot(ATT_data, aes(x = year-G, y = ATT, color = factor(G))) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "ATT over t-g",
       x = "t-g",
       y = "ATT",
       color = "Policy Enactment Year")
```


Q2
```{r}

```




Q6

When ATT(g,t) can be identified, we should have: (1) the treated group should be treated before t, that is: $t \geq g$. (2) There is still comparison group that hasn't been treated at time t, that is $t \leq 2013$

So the set $\{g,t\}$ is:

\[
\{ g, t \} \text{ such that } g \in \{1996, 2002, 2008, 2014\}, t \geq g, t \leq 2013
\]


The estimand is:
\[
ATT(g,t) = E[Y_{igt} | G_i = g, t] - E[Y_{ijt} | G_j > t]
\]

The assumption: (1) Parallel trends assumption. (2) No anticipation effects.

In the TWFE model, the comparisons are 